---
title: "Euro 2021 Player Stats Analysis"
author: "Written by Lucas Okwudishu on 06/18/2021"
output: 
    html_document:
        theme: flatly
        highlight: tango
        toc_float: true
        code_folding: show
    github_document:
        keep_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## __Introduction__

While watching a particular game in the ongoing Euro 2021 competition, a discussion between commentators about the average age of a particular team really got me curious. I immediately began to wonder what the age distribution of players on teams at the competition was like. I decided to go online and search for data on player ages, which in-turn led to an entire data analysis project. 

In this project, I'll be analyzing data such as age, number of caps, number of goals, etc for players at the Euro 2021 competition.

## __Scraping The Data__

I decided to try something new this time by web scraping the data, as opposed to just downloading a ready made dataset. This was my first time trying web scraping so I'm not able to give detailed explanations on the web scraping process. However 2 helpful resources that do just that are the [UC Business Analytics R Programming Guide](http://uc-r.github.io/scraping) and [Dataslice](https://www.youtube.com/watch?v=v8Yh_4oE-Fs).

To scrape the data, I made use of 2 popular **R** libraries, **`tidyverse`** and **`rvest`**.


```{r}
# Load libraries
library(tidyverse)
library(rvest)
library(patchwork)
```

---

Next I scraped [wikipedia](https://en.wikipedia.org/wiki/UEFA_Euro_2020_squads) to get data about participating countries using the following steps:

+ Step 1: Scrape the webpage for participating countries.

+ Step 2: Scrape the webpage names of team coaches.

+ Step 3: Scrape the webpage for data tables for all 24 participating countries. Data Tables included data on player names, age,  positions (goal keepers, defenders, etc.), number of international caps, number of international goals, and players current club.

+ Step 4: Clean! Clean!! Clean!!! Clean, format and merge datasets.

### __Step 1: Scrape List of Participating Countries__

Using functions from the **`rvest`** library, I extracted a list of **h3** headers from the web page, then filtered for the first 24 **h3** headers which were the names of participating countries.


```{r}
# Wikipedia page url
url <- "https://en.wikipedia.org/wiki/UEFA_Euro_2020_squads"

# Covert url into html object
page <- read_html(url)

# Extract list of participating countries
countries_list <- page %>% 
  html_nodes("h3") %>% 
  html_text()

countries_list <- countries_list[1:24]
countries_list
```


---

### __Step 2: Scrape Names of Team Coaches__

On the web page, a new paragraph starts right after the **h3** headers. Some of these paragraphs (html tag "p") included the names of team coaches.

I extracted all paragraphs on the webpage, added them to a dataframe, filtered for all rows that included the string "Manager", then finally removed the string "Manager: " using **`stringr`**, thus leaving only the names of the coaches.


```{r}
# Get names of coaches
coach_names <- page %>% 
  html_nodes("p") %>% 
  html_text() %>% 
  str_split("\n") %>% 
  unlist() %>% 
  tibble() %>% 
  rename(coach = ".") %>% 
  filter(grepl("Manager", coach)) %>% 
  mutate(coach = coach %>% str_remove_all("Manager: "),
         coach = coach %>% str_trim(side = "both"))

coach_names %>% head()
```

---

### __Step 3: Scrape Data Tables for Participating Countries__

On the webpage, there are 24 data tables which I scraped next. I did so by first extracting the html tag "table", then once again filtering for the first 24 tables into a list. I show the code below and display the first table in the list.


```{r}
# Get data tables
table_list <- html_nodes(page, "table")

table_list <- table_list %>% 
  .[1:24] %>% 
  html_table(fill = TRUE)

table_list[[1]] %>% head()
```

---

### __Step 4: Clean! Clean!! Clean!!!__

The first thing I did here was to create a dataframe of the countries and coaches.

```{r}
# Dataframe of country & coach
country_coach_tbl <- tibble(country = countries_list) %>% 
  bind_cols(coach_names)

country_coach_tbl %>% head()
```

---

The second thing I did here was create a custom function to merge the country and coach columns from above to the corresponding data table for that country. Next I applied that function each of the 24 data tables, while also during a **`rbind`** to merge all 24 tables.

```{r}
# Custom function to merge data table with corresponding country and coach
func_add_country_coach <- function(table_list_no = 1, coach_country_no = 1){
  
  table_list[[table_list_no]] %>% 
    mutate(country  = country_coach_tbl[coach_country_no,]$country,
           coach    = country_coach_tbl[coach_country_no,]$coach)
}

# Apply custom function from above to all 24 tables while binding rows
euros_tbl <- tibble(
  rbind(
  func_add_country_coach(1, 1), func_add_country_coach(2, 2),
  func_add_country_coach(3, 3), func_add_country_coach(4, 4),
  func_add_country_coach(5, 5), func_add_country_coach(6, 6),
  func_add_country_coach(7, 7), func_add_country_coach(8, 8),
  func_add_country_coach(9, 9), func_add_country_coach(10, 10),
  func_add_country_coach(11, 11), func_add_country_coach(12, 12),
  func_add_country_coach(13, 13), func_add_country_coach(14, 14),
  func_add_country_coach(15, 15), func_add_country_coach(16, 16),
  func_add_country_coach(17, 17), func_add_country_coach(18, 18),
  func_add_country_coach(19, 19), func_add_country_coach(20, 20),
  func_add_country_coach(21, 21), func_add_country_coach(22, 22),
  func_add_country_coach(23, 23), func_add_country_coach(24, 24)
  
  )
)

euros_tbl %>% head()
```

---

Finally I did some additional cleaning/formatting including extracting the players age, added a column to indicate the players position, added another column to indicate if a player is the team captain or not.

```{r}
remove <- c("captain|\\(|\\)")

# Extract age
final_euros_tbl <- euros_tbl %>% 
    setNames(names(.) %>% str_to_lower()) %>% 
    rename(dob = `date of birth (age)`) %>% 
    separate(col = dob, into = c("col1", "col2", "col3", "col4", "col5"), sep = " ") %>% 
    select(-c("col1", "col2", "col3", "col4")) %>% 
    rename(age = col5) %>% 
    mutate(age = age %>% str_remove_all("\\)") %>% as.numeric) %>% 
    
    # Add position column
    mutate(position = case_when(
    str_detect(pos., "GK") ~ "Goalkeepers",
    str_detect(pos., "DF") ~ "Defenders",
    str_detect(pos., "MF") ~ "Midfielders",
    str_detect(pos., "FW") ~ "Forwards",
    )) %>% 
    mutate(position = position %>% fct_relevel("Goalkeepers", "Defenders", "Midfielders", "Forwards")) %>% 
    
    # Add captain indicator column
    mutate(captain = case_when(
        str_detect(player, "(captain)") ~ "Yes",
        TRUE ~ "No"
    )) %>% 
    
    # Remove the string "(captain)" from the "player" column
    mutate(player = player %>% str_remove_all(remove)) %>% 
    mutate(player_country = str_glue("{player} ({country})"))

final_euros_tbl %>% head()
```

At this stage, I'm done with all the cleaning and formatting and have a clean dataset to analyze. Before I proceed to any analysis, I checked to see if there were any NA values.

```{r}
# Check For NAs
final_euros_tbl %>% 
    sapply(function(x) sum(is.na(x)))
```

Great! No NA values.

---

## __Analyzing The Data__

I had several questions I was curious about such as: 

+ What player has the most caps for he's country?

+ What player has the most goals for he's country?

+ Who are the youngest/oldest players in the tournament?

+ What is the age distribution ob players in the tournament?

I then proceeded to get the answers to these questions from the data and visualize using barcharts.

### __What players have the most caps for their country?__

Note: caps refers to the number of international games a player has played for he's country.

```{r, out.width="80%"}
# Function to create bar charts
func_plot_bar_chart <- function(data, x_var, y_var, label_var){
    
    x_var_expr <- rlang::enquo(x_var)
    y_var_expr <- rlang::enquo(y_var)
    label_var_expr <- rlang::enquo(label_var)
    
    data %>% 
    ggplot(aes(!!x_var_expr, !!y_var_expr))+
    geom_col(fill = "#669cd1", width = 0.7)+
    geom_text(aes(label = !!label_var_expr), hjust = 1, size = 3.5)+
    theme_minimal()+
    theme(axis.text.x = element_text(size = 10, colour = "black"),
                   axis.text.y = element_text(size = 10, color = "black"))
}

# Top 10 capped players
final_euros_tbl %>% 
    arrange(desc(caps)) %>% 
    slice_max(order_by = caps, n = 10) %>% 
    mutate(player_country = player_country %>% fct_reorder(caps)) %>% 
    func_plot_bar_chart(x_var = caps, y_var = player_country, label_var = caps)+
    labs(title = "Top 10 Capped Players", y = "", x = "\nNo. of Caps")
```

---

### __Who are the highest capped players under 30?__


```{r, out.width="80%"}
# Top 10 capped players under 30
final_euros_tbl %>% 
    filter(age < 30) %>% 
    arrange(desc(caps)) %>% 
    slice_max(order_by = caps, n = 10) %>% 
    mutate(player_country = player_country %>% fct_reorder(caps)) %>% 
    func_plot_bar_chart(x_var = caps, y_var = player_country, label_var = caps)+
    labs(title = "Top 10 Capped Players Under 30", y = "", x = "\nNo. of Caps")
```

---

### __Who are the highest capped players under 25?__

```{r, out.width="80%"}
final_euros_tbl %>% 
    filter(age < 25) %>% 
    arrange(desc(caps)) %>% 
    slice_max(order_by = caps, n = 10) %>% 
    mutate(player_country = player_country %>% fct_reorder(caps)) %>% 
    func_plot_bar_chart(x_var = caps, y_var = player_country, label_var = caps)+
    labs(title = "Top 10 Capped Players Under 25", y = "", x = "\nNo. of Caps")
```

---

### __What players have the most goals for their country?__

```{r, out.width="80%"}
# Top 10 goal scorers
final_euros_tbl %>% 
    arrange(desc(goals)) %>% 
    slice_max(order_by = goals, n = 10) %>% 
    mutate(player_country = player_country %>% fct_reorder(goals)) %>% 
    func_plot_bar_chart(x_var = goals, y_var = player_country, label_var = goals)+
    labs(title = "Top 10 Goal Scorers", y = "", x = "\nNo. of Goals")
```

Understandably, players (forwards especially) who have a high number of caps will also have a higher number of international goals than players who may just be early or mid-way through their international careers. Thus I wanted to look at the ratio of goals to caps for players with at least 5 international caps, 10 international caps, 50 international caps, and 100 international caps.

```{r, out.width="80%"}
# Add caps bin to dataset 
goals_caps_tbl <- final_euros_tbl%>% 
    mutate(caps_bin = case_when(
        caps >= 5  & caps <= 10  ~ "5 - 10 caps",
        caps >= 10 & caps <= 50  ~ "11 - 50 caps",
        caps >= 50 & caps <= 100 ~ "51 - 100 caps",
        caps >= 100 ~ "Above 100 caps",
        TRUE ~ "Exclude")) %>% 
    mutate(caps_bin = caps_bin %>% fct_relevel("5 - 10 caps", "11 - 50 caps",
                                               "51 - 100 caps", "Above 100 caps",
                                               "Exclude")) %>% 
    filter(caps_bin != "Exclude") %>% 
    mutate(goal_to_cap_ratio = goals / caps) %>% 
    mutate(label_text = str_glue("{goals} | {caps} | {goal_to_cap_ratio %>% scales::percent(accuracy = 1)}"))

# Function to plot goals to caps ratio
func_plot_goal_cap_ratio <- function(data){
    
    data %>% 
        mutate(goal_to_cap_ratio = goals / caps) %>%
        arrange(desc(goal_to_cap_ratio)) %>% 
        mutate(player_country = player_country %>% fct_reorder(goal_to_cap_ratio)) %>% 
        slice_max(order_by = goal_to_cap_ratio, n = 10) 
}

# Goals to caps plot: 5 - 10 caps
goals_caps_tbl %>% 
    filter(caps_bin == "5 - 10 caps") %>% 
    func_plot_goal_cap_ratio() %>% 
    func_plot_bar_chart(x_var = goal_to_cap_ratio, y_var = player_country, label_var = label_text)+
    scale_x_continuous(labels = scales::percent_format(accuracy = 1))+
    labs(title = "Top 10 Players: Goals to Caps Ratio (5 - 10 Caps)", subtitle = "Goals | Caps | Ratio",
         x = "\nGoals to Caps Ratio", y = "")

# Goals to caps plot: 10 - 50 caps
goals_caps_tbl %>% 
    filter(caps_bin == "11 - 50 caps") %>% 
    func_plot_goal_cap_ratio() %>% 
    func_plot_bar_chart(x_var = goal_to_cap_ratio, y_var = player_country, label_var = label_text)+
    scale_x_continuous(labels = scales::percent_format(accuracy = 1))+
    labs(title = "Top 10 Players: Goals to Caps Ratio (11 - 50 Caps)", subtitle = "Goals | Caps | Ratio",
         x = "\nGoals to Caps Ratio", y = "")

# Goals to caps plot: 50 - 100 caps
goals_caps_tbl %>% 
    filter(caps_bin == "51 - 100 caps") %>% 
    func_plot_goal_cap_ratio() %>% 
    func_plot_bar_chart(x_var = goal_to_cap_ratio, y_var = player_country, label_var = label_text)+
    scale_x_continuous(labels = scales::percent_format(accuracy = 1))+
    labs(title = "Top 10 Players: Goals to Caps Ratio (51 - 100 Caps)", subtitle = "Goals | Caps | Ratio",
         x = "\nGoals to Caps Ratio", y = "")

# Goals to caps plot: above 100 caps
goals_caps_tbl %>% 
    filter(caps_bin == "Above 100 caps") %>% 
    func_plot_goal_cap_ratio() %>% 
    func_plot_bar_chart(x_var = goal_to_cap_ratio, y_var = player_country, label_var = label_text)+
    scale_x_continuous(labels = scales::percent_format(accuracy = 1))+
    labs(title = "Top 10 Players: Goals to Caps Ratio (Above 100 caps)", subtitle = "Goals | Caps | Ratio",
         x = "\nGoals to Caps Ratio", y = "")
```

---

### __Who are the oldest and youngest players in the tournament?__


```{r, out.width="80%"}
# Oldest players
final_euros_tbl %>% 
    arrange(desc(age)) %>% 
    slice_max(order_by = age, n = 10) %>% 
    mutate(player_country = player_country %>% fct_reorder(age)) %>% 
    func_plot_bar_chart(x_var = age, y_var = player_country, label_var = age)+
    labs(title = "10 Oldest Players", x = "\nAge", y = "",
         caption = "More than 10 value returned when multiple players share the same age")

# Youngest players
final_euros_tbl %>% 
    arrange(age) %>% 
    slice_min(order_by = age, n = 10) %>% 
    mutate(player_country = player_country %>% fct_reorder(age)) %>% 
    func_plot_bar_chart(x_var = age, y_var = player_country, label_var = age)+
    labs(title = "10 Youngest Players", x = "\nAge", y = "",
         caption = "More than 10 value returned when multiple players share the same age")
```

---

### __What is the age distribution of players like?__


```{r, out.width="80%"}
# Age distribution (all)
p1 <- final_euros_tbl %>% 
    ggplot(aes(age))+
    geom_boxplot(fill = "#669cd1")+
    theme_minimal()+
    labs(title = "Age Distribution",
         subtitle = str_glue("Min Age: {min(final_euros_tbl$age)} | Average Age: {round(mean(final_euros_tbl$age),0)} | Max Age: {max(final_euros_tbl$age)}"))

# Age distribution by position
p2 <- final_euros_tbl %>% 
    ggplot(aes(age))+
    geom_boxplot(fill = "#669cd1")+
    theme_minimal()+
    facet_wrap(~ position)+
    labs(title = "Age Distribution by Position", x = "Age")

p1 / p2
```


---

### __What clubs are most represented at the tournament?__

```{r, out.width="80%"}
# Clubs with the highest representation at the tournament
final_euros_tbl %>% 
    group_by(club) %>% 
    count() %>% 
    arrange(desc(n)) %>% 
    ungroup() %>% 
    slice(1:10) %>% 
    mutate(club = club %>% fct_reorder(n)) %>% 
    func_plot_bar_chart(x_var = n, y_var = club, label_var = n)+
    labs(title = "Top 10 Clubs Represented", x = "No. of Players", y = "")
```

---

### __What Countries have the most players at a single club?__

```{r, out.width="80%"}
# Countries with the most players at a single club
final_euros_tbl %>% 
    group_by(country, club) %>% 
    count() %>% 
    arrange(desc(n)) %>% 
    ungroup() %>% 
    filter(n >= 5) %>% 
    mutate(country_club = str_glue("{country} ({club})")) %>% 
    mutate(country_club = country_club %>% fct_reorder(n)) %>%  
    func_plot_bar_chart(x_var = n, y_var = country_club, label_var = n)+
    labs(title = "Countries With the Most Players at a Single Club", x = "No of Players", y = "")
```


---

## __Wrapping Up__

In this project, I scrapped Euro 2021 data from [wikipedia](https://en.wikipedia.org/wiki/UEFA_Euro_2020_squads), then proceeded to answer some questions around player stats, using bar charts and boxplots to visualize the data.
